\documentclass[12pt]{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{gensymb}
\usepackage[round, authoryear]{natbib} 
\usepackage[a4paper, margin=2cm]{geometry}
\usepackage{setspace}

\singlespacing
\bibliographystyle{plainnat} 


\title{Winter 2026: DS9600 Final Project \\ \textbf {Beyond the Black Box: A Multimodal XAI Framework for Lumbar Spinal Stenosis Grading via Integrated MRI Analysis and LLM-Driven Clinical Reasoning}}
\author{Ahmed Mohamed, (Helen) Yixiu He, Yan Yi Li, Tim Nguyen}

\begin{document}
\maketitle


\section{Problem Statement}
\label{problem_statement}
In the field of medical imaging, medical practitioners deal with a rising diagnostic workload.  To mitigate this issue, many choose to extend work hours and increase productivity expectations \citep{Jing2025}. However, this promotes job dissatisfaction and burnout. In private practice, the burnout rate is 46\% for radiologists. This result strongly correlates with practitioners who took evenings, overnight and weekends calls \citep{Parikh2023}. Not just in private sector, but burn out also happens in acadmedia, where burnout overall prevalence rate is 37.4\% \citep{Higgins2021}. Even though practices provide solution to mitigate this issue such as flexible scheduling, part-time and remote work options, teleradiology services \citep{Rawson2024}, these only address the surface problems instead of targeting the root cause, which is the shortage of radiologists  \citep{Jing2025}. With the increasing number of patients but shortage of professionals to diagnose the images, it causes huge delay and decrease in quality in health care providing. Due to an increasing of already heavy work flow, combining with ineffective operation and shortage of radiologists, this leads to excessive workload, with visual and mental fatigue. These can lead to imaging delays, causing delays of diagnosis and treatment \citep{Cournane2016}. In 2025, a survey reports that 11.4\% of patient had to stop working while waiting for a result of a scan. The wait time can vary between one up to three months \citep{nanos}. 
For that issue, we are interested in building a pipeline to address Lumbar Spinal Stenosis (LSS), but the general idea is applicable for any medical imaging. In order to address this problem, researchers developed a pipeline to utilize the power of computer vision to aid medical practitioners in diagnosing imaging. In particular, there exists a pipeline to use U-Net architecture to perform ROI segmentation \citep{Silveira2025}, ConvNet in image classification and Grad-CAM to generate feature importance heatmap \citep{Selvaraju2019}. However, in a clinical survey, 56\% radiologists mentioned a lack of transparency since this pipeline utilizes black box models \citep{aravazhi2024}, there is lack of motivation behind predictions. Furthermore, they state that they prefer textual explanation \citep{gale2018producingradiologistqualityreportsinterpretable, kayser2024fooloncecontrastingtextual} , which our model comes to play. Here, we propose the use of multimodal large language model to attempt to explain the motivation behind prediction using ROI and heatmap. 

\section{Rationale and Background}
\subsection{Clinical Burden and Diagnostic Variability}
Lumbar Spinal Stenosis (LSS) is a leading cause of neurogenic claudication and a primary indication for spinal surgery in adults over 60, affecting an estimated 11\% of the general population \citep{Munakomi2023}. Magnetic Resonance Imaging (MRI) is the gold standard for diagnosis; however, the manual assessment of multiple spinal levels (L1–S1) is labor-intensive and suffers from significant inter-observer variability. Studies indicate that even expert radiologists show only “moderate” agreement (Cohen’s kappa 0.40–0.57) when grading stenosis severity \citep{AlTameemi2017, Holc2023}. This diagnostic inconsistency can lead to misaligned treatment plans, varying from unnecessary surgeries to delayed interventions. Due to these two reasons, developing a tool to support diagnosis process and improve accuracy is essential for this condition. 

\subsection{The Rise and Limitations of Deep Learning in Biology}
Recent advancements in Deep Learning (DL) have demonstrated that Convolutional Neural Networks (CNNs) can achieve expert-level accuracy in LSS classification. Models utilizing architectures like U-Net for segmentation with CNN-based classifiers have reached accuracy exceeding 0.94 \citep{Liu2025}. Despite this high performance, these models primarily function as “black boxes.” In a clinical survey by Bioinformation, 56\% of radiologists cited a lack of transparency as a major barrier to AI adoption \citep{aravazhi2024}. While visual tools like Grad-CAM (Gradient-weighted Class Activation Mapping) highlight “hotspots” of model attention, they are often criticized in medical literature for being “noisy” and anatomically imprecise, failing to distinguish between pathologically relevant areas and background noise \citep{Houssein2025, Muhammad2024}.

\subsection{The Unfeasiability of Interpretable AI}
One option for this problem is adopting interpretable models instead. In high-stake industries such as medicine and law, the use of interpretable models such as linear regression, logistic regression and decision list is favoured due to their transparency, as opposed to black box models where the content is opaque even to its developers \citep{wang2019hybridpredictivemodelinterpretable}. Many literatures encourage the pursuit for simple models allowing the understanding of motivation behind decisions, giving the stakeholders the necessary reasonings for decision making \citep{rudin2019stopexplainingblackbox}. However, also according to \cite{rudin2019stopexplainingblackbox}, it is particularly challenging, if not impossible, to utilize these models for complex tasks, in which one of them is computer vision and imaging. In this pipeline, since we exploit the power of computer vision techniques as decision making assistants, it is virtually not possible to implement simple models that can understand images in an effective manner. That leaves us one other possible solution for this conundrum.

\subsection{Explainable AI \& its “Semantic Gap”}

In this proposal, we suggest the use of explainable AI (XAI) to mitigate the opaqueness of black box models. In the current pipeline, a heatmap generated by using Grad-CAM to highlight the important area on the imaging that influences the prediction. However, there is a disconnection between pixel-level importance and clinical reasoning. A clinician does not make a diagnosis based on a heatmap; they make it based on anatomical observations such as “thecal sac compression” or “posterior disc bulging.” Research in human-computer interaction (HCI) suggests that clinicians strongly prefer textual explanations that mirror the structure of a radiology report over purely visual saliency maps (Gale et al., 2018; Kayser et al., 2024). Furthermore, visual-only explanations can lead to automation bias, where clinicians may trust a correct prediction for the wrong underlying reason (e.g., the model focused on a surgical clip rather than the spinal canal) \citep{Houssein2025}. This problem also aligns with a study by \cite{panboonyuen2026seeingisntbelievinganalysis}, where they found heatmaps generated by Grad-CAM fail to represent true motivation behind a decision. Therefore, to make heatmap more accessible and helpful to medical practitioners, textual explanation is needed, hence the adoption of Large Language Models (LLMs). 

\subsection{Identifying the Gap: The Need for Vision-Language Fusion}
Large Language Models (LLMs) are models that are trained on a large dataset of text. They are able to provide concise and easy to comprehend textual explanations. In this proposal, we are suggesting the use of Multimodal Large Language Models (MLLMs). MLLMs are trained to connect different modalities such as visual, audio and text together to provide a comprehensive presentation that connect the three sectors. Here, we are proposing the use of MR imaging, MRI segmented ROI, heatmap generated by Grad-CAM and prediction by the pipeline as inputs for the MLLMs to generate textual explanation.  
However, while LLMs have shown remarkable ability in synthesizing medical information, they lack “grounding” in raw imaging data when used in isolation, leading to “hallucinations” \citep{pal2024geminigoesmedschool, Omar2025, Yang2025} . The identified gap is the lack of a pipeline that translates quantitative, segmented anatomical features and visual attention patterns into structured, clinical-grade natural language. 
By anchoring an LLM’s reasoning in morphometric data (e.g., exact canal area measurements) and spatial overlap metrics (how much the AI’s attention overlaps with the thecal sac), our proposal creates a novel, interdisciplinary “translation layer.” This addresses the technical challenge of XAI while fulfilling the clinical requirement for evidence-based diagnostic support, ultimately moving AI from a “scoring tool” to a “transparent diagnostic partner.”

\section{Proposed ML/AI System or Solution}
This study aims to develop an interpretable deep learning framework for automated severity assessment of Lumbar spinal stenosis (LSS) using axial T2-weighted Magnetic Resonance Imaging (MRI), accompanied by a structured text-based clinical explanation of the model’s decision. The proposed framework consists of two sequential components: anatomical structure segmentation with quantitative morphometric analysis, followed by stenosis severity classification with integrated explainability.

\paragraph{Datasets}
For anatomical segmentation, we will use a publicly available lumbar spine MRI dataset from Mendeley Data (Dataset ID: k57fr854j2), derived from an anonymized clinical cohort of patients with lower back pain \citep{lumbardataset}. The dataset includes T2-weighted lumbar MRI scans and a companion annotated subset (Dataset ID: k57fr854j2.1) with expert-generated segmentation masks for key anatomical structures (e.g., intervertebral discs and spinal canal) \citep{labelImg}, which will serve as the reference standard for model training and evaluation. A curated subset (around 500 mid-sagittal slices) will be used. Also, the dataset will be split at the patient level into training/validation (80\%) and testing (20\%) sets to prevent data leakage. For stenosis severity classification, we will utilize the RSNA Lumbar Spine Degenerative Classification Dataset (n = 2,697), which provides vertebral-level stenosis grades (normal, mild, moderate, severe) \citep{Richards2026}. The segmentation model trained on the Mendeley dataset will be applied to the RSNA dataset to extract consistent anatomical regions of interest (ROIs) and quantitative features across all cases, enabling standardized input for downstream classification.

\paragraph{Phase 1: Quantitative ML Prediction Framework}
We will develop a quantitative, anatomically grounded machine learning (ML) framework for lumbar spinal stenosis (LSS) assessment from axial lumbar spine MRI. A multi-class 2D U-Net-based convolutional neural network (CNN) \citep{ronneberger2015unetconvolutionalnetworksbiomedical} will be used to segment key anatomical structures implicated in LSS, including the intervertebral disc (IVD), posterior elements (PE), thecal sac (TS), and the anterior–posterior spinal canal (AAP). A 2D architecture is selected to align with clinical practice, where stenosis grading is performed on axial slices and morphology-based systems (e.g., Schizas classification) rely on slice-level evaluation of the thecal sac. The segmentation model will be trained in a supervised manner using a composite Dice and focal loss to address class imbalance and improve boundary delineation. Performance will be evaluated using Dice Similarity Coefficient (DSC) and Intersection-over-Union (IoU) on a held-out test set. From the segmentation outputs, quantitative morphometric features will be extracted, including cross-sectional area (CSA) computed using pixel spacing from DICOM metadata. These measurements serve as interpretable anatomical biomarkers, with particular emphasis on thecal sac area reduction as a primary indicator of stenosis severity.
A region of interest (ROI) encompassing the segmented structures will then be automatically cropped and used as input to a convolutional classifier, EfficientNets \citep{tan2020efficientnetrethinkingmodelscaling}. The classifier will predict stenosis severity (none, mild, moderate, severe) along with calibrated prediction probabilities. Performance will be assessed using accuracy, sensitivity, specificity, weighted F1-score, and Cohen’s kappa to quantify agreement with expert annotations. To enhance interpretability, Grad-CAM \citep{Selvaraju2019} will be applied to generate attention maps highlighting regions that drive model predictions. These maps will be quantitatively compared with segmentation masks to compute attention-anatomy overlap (e.g., Dice overlap), with a focus on clinically critical structures such as the thecal sac. In addition, the proportion of total Grad-CAM activation within the ROI will be calculated to assess whether model attention is spatially aligned with relevant anatomy. Finally, predicted severity and confidence, quantitative anatomical measurements, attention-structure overlap metrics, and ROI activation percentages will be integrated into a structured representation.

\paragraph{Phase 2: LLM-Based Structured Explanation Framework}
To improve interpretability and clinical usability, we will develop a large language model (LLM)-based framework that generates standardized, clinically meaningful explanations grounded in quantitative imaging evidence. Structured outputs from Phase 1will be converted into templated prompts (Appendix 1). The LLM will generate concise, clinically styled explanations that explicitly link model predictions to anatomical findings. To ensure reliability and minimize unsupported claims, all outputs will be constrained using predefined prompt templates and rule-based grounding mechanisms (Appendix 1). The system will also incorporate explicit safety constraints: the LLM will only reference the provided quantitative values, will not perform independent clinical inference beyond the structured inputs, and every generated explanation will include a clear AI-assistance disclaimer indicating that the output is intended to support, rather than replace, clinical judgment.

\paragraph{Phase 3: Clinician Evaluation and Workflow Assessment}
To assess the clinical utility and workflow impact of the proposed system, a reader study will be conducted with radiologists using a controlled, crossover design. The system will be deployed in a simulated radiology workstation environment where users can review axial T2-weighted MRI scans and interact with AI outputs. Evaluation will be based on three workflow conditions: (1) standard workflow: radiologists interpret MRI scans independently and generate reports from scratch without AI assistance; (2) AI-only decision workflow: radiologists are provided with predicted stenosis severity and confidence score only. In this setting, clinicians must independently verify the AI prediction by reviewing the images and reconstructing the reasoning before producing a report. (3) AI-assisted workflow (decision + reasoning-based structured report): In this condition, the system provides explicit reasoning and a structured report draft, allowing clinicians to validate the AI’s decision pathway directly and edit the generated explanation report rather than writing reports from scratch.
System performance will be evaluated across efficiency, diagnostic accuracy, and user experience. Efficiency will be measured by the time required to complete each case and finalize the report. Diagnostic performance will be assessed using accuracy, sensitivity, specificity, and agreement with ground truth through Cohen’s kappa. User experience will be evaluated through a structured questionnaire assessing trust in AI predictions, perceived usefulness of explanations, cognitive workload, and reporting confidence.
It is expected that the AI decision only workflow will not significantly reduce the time, as clinicians must independently verify the prediction, effectively duplicating the decision-making process and then writing the report from scratch. In contrast, the AI-assisted workflow with reasoning is anticipated to provide the greatest efficiency gains by supporting both decision validation and report generation. The standard workflow is expected to remain the least efficient.

\paragraph{Risk Mitigation}
A composite reliability score will be defined by combining classification certainty (e.g., softmax probability or confidence score) with an attention–anatomy consistency measure, computed as the overlap between Grad-CAM activation maps and segmented clinically relevant regions (e.g., thecal sac and spinal canal). Predictions with low classification confidence and/or low attention overlap will be flagged as unreliable. A predefined threshold will be established to identify such cases. When this threshold is not met, the system will explicitly notify the user that the prediction may be unreliable and discourage reliance on the AI output, prompting full independent clinical review.

\section{Responsible AI: Bias, Ethics, and Equity}
\subsection{Data Bias and Population Representation}
AI models for musculoskeletal imaging are only as equitable as the data on which they are trained. A significant risk in LSS classification is representation bias. Clinical studies indicate that spinal morphology and the prevalence of degenerative changes vary significantly across different ethnic groups and body mass indices (BMI) \citep{Williams2015, Wng2023, Kalichman2009}. For instance, high-BMI patients often have lower-quality MRI scans due to signal-to-noise ratio limitations, which can lead to higher AI error rates for this subgroup. 
For the mitigation, we will implement disaggregated performance auditing, evaluating the model’s accuracy, calibration, and F1-score across different demographic strata (age, sex, and BMI) to ensure no specific group suffers from diagnostic disparity \citep{vandeSande2022, Lu2022} .

\subsection{Automation Bias and the “Right to Explanation”}
As discussed in Section 2, a primary ethical concern in clinical AI is automation bias, the tendency for over-reliance on automated suggestions \citep{Bergomi2025} . Purely visual explanations (heatmaps) can exacerbate this by allowing clinicians to subconsciously “rationalize” a model’s incorrect prediction. 
By introducing an LLM-driven textual reasoning layer, this design forces a “logic check”: if the LLM’s textual description (e.g., “narrowing at L4-L5”) contradicts the clinician’s observation of the raw MRI, the friction created by this discrepancy serves as a critical safeguard against erroneous AI-led decisions \citep{gale2018producingradiologistqualityreportsinterpretable, kayser2024fooloncecontrastingtextual}.

\subsection{Data Privacy and Governance in Multimodal Systems}
The use of MRI scans and the synthesis of clinical reports by LLMs raise significant privacy concerns. MRI DICOM files contain sensitive metadata, and LLMs, if not properly governed, can inadvertently “memorize” and leak Protected Health Information (PHI). 
To deal with it, our system will employ de-identification pipelines before data processing. Furthermore, we will utilize frozen, locally-hosted LLMs (or HIPAA-compliant API instances) to ensure that patient-specific anatomical features extracted in Stage 4 do not enter the public training pool of commercial AI providers \citep{Zhong2025}.

\subsection{Equity: Accessibility and Linguistic Inclusivity}
While advanced AI tools are often limited to high-resource urban hospitals, our system’s focus on textual explanation can actually promote health equity. By providing a clear, evidence-based reasoning summary, the tool can support non-specialist clinicians (e.g., family doctors in rural Ontario) in making more accurate triage decisions, thereby reducing unnecessary referrals to overburdened specialists \citep{Vergara2025}. 
For future equity goals, we acknowledge the risk of linguistic bias in LLMs, which primarily perform best in English. To ensure equity for Canada’s diverse population, future iterations must validate the LLM’s explanatory performance in French and other widely spoken languages to prevent “interpretability gaps” for non-English speaking practitioners.

\section{Stakeholders and Implementation Context}
\subsection{Stakeholders}
Radiologists are the primary users of the system. They will benefit from automated severity assessments accompanied by structured, reasoning-based reports. The system is designed as a decision-support tool, providing an initial diagnostic report that the radiologist can review and finalize.

Patients are indirect beneficiaries. While they do not access the AI directly, radiologists can use the reasoning-based reports to create clear, patient-friendly summaries. By supporting consistent severity assessments and highlighting key anatomical findings, the system helps ensure reliable, understandable, and standardized communication of lumbar spinal stenosis results, enhancing patient comprehension and shared decision-making.

Hospital IT Departments are essential for deployment and maintenance, ensuring smooth integration with PACS/DICOM systems, secure handling of patient data, and compliance with hospital IT infrastructure and regulatory standards.

\subsection{Implementation}
The system will be deployed as a supporting tool, not a replacement for clinical judgment. It provides initial results and a reasoning-based draft report that highlights the anatomical basis for the AI’s prediction, such as reductions in thecal sac or nerve root areas at specific vertebral levels. Radiologists can quickly validate the AI’s assessment by focusing on relevant regions, rather than performing a full conventional review, and edit the report as needed. By acting as an editable draft, similar to reviewing a trainee’s report, the system saves time in both validating AI decisions and generating final reports, while maintaining full human oversight and compliance with regulatory guidelines.
Radiologists and IT staff will be involved during co-design, and evaluation phases to ensure usability and integration with existing workflows. 

Feasibility: Integration with existing PACS/DICOM systems, compatibility with hospital IT systems and security protocols.

Acceptability and Uptake: Clinical adoption will depend on demonstrable time savings, consistency of severity grading, interpretability of AI outputs, and alignment with regulatory guidelines (e.g., Health Canada’s Software as a Medical Device framework).


\section{Innovation \& Contribution}
The proposed framework introduces a novel method for the automated detection and classification of LSS severity through two primary innovations: (1) A Modular Vision-to-Text Diagnostic Pipeline: We propose a staged architecture that mimics clinical diagnostic logic: a 2D U-Net first performs precise anatomical segmentation to extract quantitative morphometrics and isolate Region of Interest (ROI), which is then passed to a specialized classifier for severity grading. This is integrated with a Zero-Shot NLP engine that organizes these measurements and visual attention features (Grad-CAM) into a structured reasoning report. By informing the physician of the exact evidence used to reach a decision, this framework transforms the AI from a "black box" into a transparent assistant, drastically reducing the time required for clinical validation. (2) Automated Reliability Auditing via Spatial Attention Overlap. To ensure the model focuses on relevant pathology, we introduce a "Reliability Score" that provides a quantitative check of the model's logic. This score mathematically measures the spatial overlap between the model’s visual attention (Grad-CAM) and the anatomical segmentation masks it generated. By verifying that the model's focus is physically tethered to the relevant anatomy, rather than image noise or artifacts, we establish a verifiable standard for trust and interpretability in spinal imaging AI.



\clearpage
\appendix
\section{Appendix}
\subsection{Appendix 1: Prompt Template Design}
\paragraph{Layer 1: System instruction}
\textit{You are an AI assistant generating clinical explanations for lumbar spinal stenosis (LSS).Please use only the provided structured inputs and Ensure all statements are supported by quantitative evidence. Do not infer or assume information not explicitly given and do not provide diagnosis beyond the predicted class.}
\paragraph{Layer 2: Structured data input}
\textit{Here is the structured inputs:
Predicted severity: Moderate
Confidence: 0.87
Anatomical measurements:
- Thecal sac area: 62 mm²
Attention overlap (Dice):
- Thecal sac: 0.71
- Intervertebral disc: 0.32
- Posterior elements: 0.28
ROI activation: 89\%}
\paragraph{Layer 3: Reasoning instructions}
\textit{Please provide severity explanation by prioritizing thecal sac measurements, using attention overlap to support whether the model focused on relevant anatomy. If ROI activation < 50\%, mention potential attention dispersion. If confidence < 0.6, include uncertainty language.}
\paragraph{Layer 4: Output format}
\textit{Please generate the output by including predicted severity and confidence as severity assessment, key measurements and their implication as anatomical evidence, alignment between attention and anatomy as model attention, summary, and AI-generated disclaimer.}

\clearpage

\bibliography{references.bib}

\end{document}
